---
title: "lesson_DMFR_11-21-22"
author: "tara"
date: "11/20/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Set up ###

## Load relevant packages ##
```{r, warning=FALSE, echo=FALSE}
library(tidyverse)
```

## Set working directory ##
```{r}
setwd("~/Documents/R-Github/DMFR_Users")
```

## Read in data
```{r}
tidepool_dat <- read.csv("data/tidepool_data.csv", header =TRUE)
```

**Inspect the data**
```{r}

```

# **Core dplyr functions**
select() #subsets/reorders columns
filter() #subsets by row, works similarly to subset()
arrange() #sorts by row
mutate() #creates new column(s) based on formula(s)
summarize() #aggregates data by group(s) based on formula(s)

**Examples of how these functions are used:**
# The first argument is always the dataframe.
# NOTE: output below will print to the console, but it can be stored as per usual <-

**select**
```{r}

```

**filter**
```{r}

```

**arrange** Sort one column by values of one or more other columns
```{r}

```

**mutate**  
#mutate is similar to tidepool_dat$newcolumnname <- some_vectorized_operation, but you can create multiple columns at once
```{r}
#Create new columns with mutate
#new columns can be based on columns created in the same call (e.g. musselsx2 get used to make snailsplusmusselsx2)

# notice how none of our changes are saved unless we assign the operation to a new object.

## Mutate across

## Mutate with ifelse ###
## uses values of one column to create another column. 

#that does not account for a situation where mussels = barnacles

# nested ifelse

#ifelse with And/or statements
# let's say you had a species that only liked to forage on mussel beds at high tide
# create a new column that categorizes different observations as good forage or not good forage. 

```
**find & replace with gsub** 
```{r}
# edit a column using mutate
# gsub - a variant of grep, is like find replace and can be used 


```

**group and summarize data by column**
summarize is similar to aggregate() but you can create multiple aggregated columns at once
```{r}

#this calculates these values for the whole dataset though, which is not real useful

#to calculate based on particular a variable, you first need to specify 'grouping' variables

#when you apply the same code to the grouped dataframe, you now get a value for each site

#group by more than one variable at a time. 


#Try to avoid creating new summary columns that have the same name as a previous column


#summarize multiple columns at the same time with "summarize_at"


#useful tip that isn't in the help files:
#mutate also acts separately on grouped data, for example:
#scale data across all sites

#scale data within sites

#confirm this using summarize


```

**piping**
#The %>% operator in dplyr allows you to pass the output of a function as the 
first argument (input dataframe) for the next function, e.g.
**filter(tidepool_dat, Site=="A") is the same as tidepool_dat %>% filter(Site=="A")**
It eliminates the need to store lots of intermediates when manipulating data.
```{r}
#For example, say you wanted to do the following series of steps.

#Rather than storing 4 intermediate dataframes, you can string the operations together like so:

#The output dataframe of the previous line become the input dataframe to the next line.

#the backwards pipe! (to permanently modify the orginal dataframe, for two or more pipe chains)
#save the original (for the purposes of this exercise)
#
 # observe output - low tide is back! because the original dataframe was not permanently modified. 

# to permanently modify the original dataframe, we must assign it to a new object that has the same name.

#return tidepool2 to the original by recreating it from tidepool_dat

# OR if there are chains of more than one pipe, we can use the backwards pipe to permanently modify the original dataframe 


```

**unite, separate, rename**  
to make new variables. 
```{r}
# unite

# separate & rename (with backwards pipe!)


```

**join**
joining tables (similar to merge())
join functions are useful when working with relational tables/databases.
```{r}
## create an example dataset ###
#filter the dataset to include sites A & C only. Then add a column called "df" which signifies the dataframe it belongs to.

#filter the dataset to include sites B & C only. This time we're using %in%, but you could also use !="A" for the same result.Then add a column called "df" which signifies the dataframe it belongs to.

#### LEFT JOIN ### - the dataframe on the left will overwrite the dataframe on the right.
#if you don't supply a 'by' argument, it will join automatically based on matching column names.
#it is possible to match by multiple columns, e.g. by=c("location","year")
##if the columns you want to match don't have exactly the same name, specify e.g. by=c("location"="LOCATION")

### RIGHT JOIN ### - the dataframe on the right will overwrite the dataframe on the left. 


## FULL JOIN ### - neither dataframes will overight. columns that have duplicate rows will retain both. 


```

**pivot** 
Convert between long format and wide format data frames. 
These functions can also be piped with dplyr functions.
```{r}


####pivot_longer### 
#  rather than have separate columns for mussels, snails, barnacles, we can put all the numeric values into a column called 'Adundance' and create a new categorical column called 'Species', which has levels mussels, snails, barnacles. This is a conversion to long format.



####pivot_wider#####
# let's change td_long back to wide format. 


```
dplyr/tidyr functions have a tendency to convert things to 'tibbles' without asking, which sometimes aren't compatible with other code you've written for data frames. To avoid this, I will end piped sequences with %>% as.data.frame()

grouping also sometimes screws things up down the line, so it's often not a bad idea to %>% ungroup() after you group and perform operations.


**working with factor levels** from the "forcats" package
```{r}
# relabel factors with fct_relabel
# can be used standalone or within mutate. 
# standalone

# if the column you are trying to relabel is not of class factor, you must use quotes, or change it to a factor first.

# change it back with mutate


# combine factor levels with fct_collapse
# combine sites B and C into one site. 


```
