---
title: "DPLYR for DMF R USERS GROUP"
author: "Tara Dolan"
date: "11/18/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Set up ###

## Load relevant packages ##
```{r, warning=FALSE, echo=FALSE}
library(tidyverse)
```

## Set working directory ##
```{r}
setwd("~/Documents/R-Github/DMFR_Users")
```

## Read in data
```{r}
tidepool_dat <- read.csv("data/tidepool_data.csv", header =TRUE)
```

```{r}
dim(tidepool_dat) #print the dimensions of the data frame (rows columns)
head(tidepool_dat, 3) # show the first three rows, or use tails(tidepool_dat, x) to show the last x number of rows
View(tidepool_dat) #View the file in a separate window.
```

# **Core dplyr functions**
select() #subsets/reorders columns
filter() #subsets by row, works similarly to subset()
arrange() #sorts by row
mutate() #creates new column(s) based on formula(s)
summarize() #aggregates data by group(s) based on formula(s)

**Examples of how these functions are used:**
# The first argument is always the dataframe.
# NOTE: output below will print to the console, but it can be stored as per usual <-

**select**
```{r}
select(tidepool_dat, site, tide.height, mussels) #subset columns site, tide.height, mussels
select(tidepool_dat, -mussels) #subset all columns except mussels
select(tidepool_dat, mussels:barnacles) #subset columns mussels through barnacles
```

**filter**
```{r}
filter(tidepool_dat, site=="A") #select only rows where site=="A"

### AND, OR, %in% #######
filter(tidepool_dat, site=="A" | site =="B" & tide.height == "low")  #(A, OR (B & low tide))
filter(tidepool_dat, site=="A" & tide.height == "low" | site =="B" )  #(A & low tide), OR B)
filter(tidepool_dat, (site=="A" | site =="B") & tide.height == "low") # (A OR B) & low tide
filter(tidepool_dat, tide.height == "low" & site %in% c("A","B")) # (A OR B) & low tide
filter(tidepool_dat, tide.height != "high" & !site %in% c("A","B")) # site C low tide

nope_sites <-c("A","B") #let's say these sites have suspicious data and you want to filter them from the dataset
td_clean <- filter(tidepool_dat, !site %in% nope_sites) # create new "clean" dataset without the suspicious sites. 

```

**arrange** Sort one column by values of one or more other columns
```{r}
arrange(tidepool_dat, site, mussels) #arrange rows by site, then mussels
arrange(tidepool_dat, site, desc(mussels)) #use desc() to arrange in descending order
```

**mutate**  
#mutate is similar to tidepool_dat$newcolumnname <- some_vectorized_operation, but you can create multiple columns at once
```{r}
#Create new columns with mutate
#new columns can be based on columns created in the same call (e.g. musselsx2 get used to make snailsplusmusselsx2)
mutate(tidepool_dat, musselsx2=mussels*2, logbarnacles=log(barnacles),snailsplusmusselsx2=snails+musselsx2) 
# notice how none of our changes are saved unless we assign the operation to a new object.

## Mutate across
mutate(tidepool_dat, across(4:6, ~ .x*2))
tide1 <-mutate(tidepool_dat, across(mussels:barnacles, as.character))
rapply(tide1, class)
tide2 <-mutate(tidepool_dat, across(c('mussels', 'barnacles'), as.character))
rapply(tide2, class)

## Mutate with ifelse ###
## uses values of one column to create another column. 
mutate(tidepool_dat, dominant_spp = ifelse(mussels > barnacles, "mussels", "barnacles"))
#that does not account for a situation where mussels = barnacles

# nested ifelse
mutate(tidepool_dat, dominant_spp = ifelse(mussels > barnacles, "mussels", 
                                           ifelse(mussels == barnacles,"equal","barnacles")))

#ifelse with And/or statements
# let's say you had a species that only liked to forage on mussel beds at high tide
# create a new column that categorizes different observations as good forage or not good forage. 
mutate(tidepool_dat, forage_hab = ifelse(tide.height=="high" & mussels > barnacles, "good forage","suboptimal forage"))

```
**find & replace with gsub** 
```{r}
# edit a column using mutate
# gsub - a variant of grep, is like find replace and can be used 
mutate(tidepool_dat, tide.height=gsub(tide.height,"high","HIGH", ignore.case = FALSE))

```

**group and summarize data by column**
summarize is similar to aggregate() but you can create multiple aggregated columns at once
```{r}
summarize(tidepool_dat, mean=mean(mussels), n=n(), sd=sd(mussels), se=sd/sqrt(n))
#this calculates these values for the whole dataset though, which is not real useful

#to calculate based on particular a variable, you first need to specify 'grouping' variables
td.grouped <- group_by(tidepool_dat, site) #grouping here by site
summarize(td.grouped, mean=mean(mussels), n=n(), sd=sd(mussels), se=sd/sqrt(n))
#when you apply the same code to the grouped dataframe, you now get a value for each site

#group by more than one variable at a time. 
td.grouped2 <- group_by(tidepool_dat, site, tide.height)
summarize(td.grouped2, mean=mean(mussels), n=n(), sd=sd(mussels), se=sd/sqrt(n))

summarize(td.grouped2, mean=mean(mussels), n=n(), sd=sd(mussels), se=sd/sqrt(n), .groups="keep") #same grouping structure as the data

#options(dplyr.summarise.inform = FALSE)
#options(dplyr.summarise.inform = TRUE)

#Try to avoid creating new summary columns that have the same name as a previous column
summarize(td.grouped2, mussels=mean(mussels), n=n(), sd=sd(mussels), se=sd/sqrt(n), .groups="keep")


#summarize multiple columns at the same time with "summarize_at"
summarize_at(td.grouped2, vars(mussels:barnacles), list(m=mean, sd=sd))


#useful tip that isn't in the help files:
#mutate also acts separately on grouped data, for example:
#scale data across all sites
mutate(tidepool_dat, mussels_scale=scale(mussels))
#scale data within sites
mutate(td.grouped, mussels_scale=scale(mussels))
#confirm this using summarize
summarize(td.grouped, mussels_scale=scale(mussels), .groups="keep")

```

**piping**
#The %>% operator in dplyr allows you to pass the output of a function as the 
first argument (input dataframe) for the next function, e.g.
**filter(tidepool_dat, Site=="A") is the same as tidepool_dat %>% filter(Site=="A")**
It eliminates the need to store lots of intermediates when manipulating data.
```{r}
#For example, say you wanted to do the following series of steps.
data.a <- group_by(tidepool_dat, site, tide.height)
data.b <- summarize(data.a, mean=mean(mussels), n=n())
data.c <- filter(data.b, tide.height=="high")
data.d <- arrange(data.c, mean)
data.d

#Rather than storing 4 intermediate dataframes, you can string the operations together like so:
data2 <- tidepool_dat %>%
  group_by(site, tide.height) %>%
  summarize(mean=mean(mussels), n=n()) %>%
  filter(tide.height=="high") %>%
  arrange(mean)
data2
#The output dataframe of the previous line become the input dataframe to the next line.

#the backwards pipe! (to permanently modify the orginal dataframe, for two or more pipe chains)
#save the original (for the purposes of this exercise)
tidepool2 <-tidepool_dat

tidepool2 %>% filter(tide.height=="high") %>% select(-mussels, -barnacles) #observe output - low tide is gone
tidepool2 # observe output - low tide is back! because the original dataframe was not permanently modified. 

# to permanently modify the original dataframe, we must assign it to a new object that has the same name.
tidepool2 <- tidepool2 %>% filter(tide.height=="high") %>% select(-mussels, -barnacles)
tidepool2

#return tidepool2 to the original by recreating it from tidepool_dat
tidepool2 <-tidepool_dat

# OR if there are chains of more than one pipe, we can use the backwards pipe to permanently modify the original dataframe 
tidepool2 %<>% filter(tide.height=="high")%>% select(-mussels, -barnacles)
tidepool2

```

**unite, separate, rename**  
to make new variables. 
```{r}
# unite
td2 <-tidepool_dat %>% unite("siterep", site, replicate, sep="_", remove=TRUE)
td2
# separate & rename (with backwards pipe!)
td2 %<>% separate(siterep, into=c("site", "replicate"), remove=T) %>% rename(Site=site)
td2

```

**join**
joining tables (similar to merge())
join functions are useful when working with relational tables/databases.
```{r}
## create an example dataset ###
#filter the dataset to include sites A & C only. Then add a column called "df" which signifies the dataframe it belongs to.
tidepool_AC <- filter(tidepool_dat, site !="B") %>% mutate(df="tidepool_AC")

#filter the dataset to include sites B & C only. This time we're using %in%, but you could also use !="A" for the same result.Then add a column called "df" which signifies the dataframe it belongs to.
tidepool_BC <-filter(tidepool_dat, site %in% c("B", "C")) %>% mutate(df2="tidepool_BC")

#### LEFT JOIN ### - the dataframe on the left will overwrite the dataframe on the right.
tidepool2 <-left_join(tidepool_AC, tidepool_BC)
#if you don't supply a 'by' argument, it will join automatically based on matching column names.
tidepool2 <-left_join(tidepool_AC, tidepool_BC, by=c("site"))
head(tidepool2)
#it is possible to match by multiple columns, e.g. by=c("location","year")
##if the columns you want to match don't have exactly the same name, specify e.g. by=c("location"="LOCATION")

### RIGHT JOIN ### - the dataframe on the right will overwrite the dataframe on the left. 
tidepool2<-right_join(tidepool_AC, tidepool_BC)

## FULL JOIN ### - neither dataframes will overight. columns that have duplicate rows will retain both. 
tidepool2<-full_join(tidepool_AC, tidepool_BC)

```

**pivot** 
Convert between long format and wide format data frames. 
These functions can also be piped with dplyr functions.
```{r}
tidepool_dat #is a wide format data frame. 

####pivot_longer### 
#  rather than have separate columns for mussels, snails, barnacles, we can put all the numeric values into a column called 'Adundance' and create a new categorical column called 'Species', which has levels mussels, snails, barnacles. This is a conversion to long format.
td_long <- pivot_longer(tidepool_dat, names_to= "Species", values_to ="Abundance", mussels:barnacles)


####pivot_wider#####
# let's change td_long back to wide format. 
(td_wide <-pivot_wider(td_long, names_from = "Species", values_from = "Abundance"))

```
dplyr/tidyr functions have a tendency to convert things to 'tibbles' without asking, which sometimes aren't compatible with other code you've written for data frames. To avoid this, I will end piped sequences with %>% as.data.frame()

grouping also sometimes screws things up down the line, so it's often not a bad idea to %>% ungroup() after you group and perform operations.


**working with factor levels** from the "forcats" package
```{r}
# relabel factors with fct_relabel
# can be used standalone or within mutate. 
# standalone
levels(tidepool_dat$site)
unique(tidepool_dat$site)
class(tidepool_dat$site)
# if the column you are trying to relabel is not of class factor, you must use quotes, or change it to a factor first.
tidepool_dat$site <- fct_recode(tidepool_dat$site, "1"="A", "2"="B", "3"="C")
levels(tidepool_dat$site)
# change it back with mutate
tidepool_dat <-mutate(tidepool_dat, site=fct_recode(site, A="1", B="2", C="3"))
levels(tidepool_dat$site)

# combine factor levels with fct_collapse
# combine sites B and C into one site. 
(tidepool2 <-mutate(tidepool_dat, site=fct_collapse(site, "B"=c("B","C"))))

```
