---
title: "lesson_DMFR_11-21-22"
author: "tara"
date: "11/20/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Set up ###

## Load relevant packages ##
```{r, warning=FALSE, echo=FALSE}
library(tidyverse)
```

## Set working directory ##
```{r}
setwd("~/Documents/R-Github/DMFR_Users")
```

## Read in data
```{r}
td <- read.csv("data/tidepool_data.csv", header =TRUE)
```

**Inspect the data**
```{r}
head(td)
```

# **Core dplyr functions**
select() #subsets/reorders columns
filter() #subsets by row, works similarly to subset()
arrange() #sorts by row
mutate() #creates new column(s) based on formula(s)
summarize() #aggregates data by group(s) based on formula(s)

**Examples of how these functions are used:**
# The first argument is always the dataframe.
# NOTE: output below will print to the console, but it can be stored as per usual <-

**select**
```{r}

select(td, -mussels)
select(td, mussels:barnacles)
```

**filter**
```{r}
filter(td, site=="A" | site=="B" & tide.height=="low") #(A, OR (B & low tide))
#(A & low tide), OR B)
filter(td, site=="A" & tide.height == "low" | site =="B" )
filter(td, (site=="A" | site =="B") & tide.height == "low") # (A OR B) & low tide

# (A OR B) & low tide %in% 
filter(td, tide.height=="low", site %in% c("A", "B"))
# site C low tide
filter(td, tide.height != "high" & !site %in% c("A","B"))

nope_site <-c("A", "B")
td_clean <- filter(td, !site %in% nope_sites) # create new "clean" dataset without the suspicious sites. 
```

**arrange** Sort one column by values of one or more other columns
```{r}
arrange(td, site, mussels) #arrange rows by site, then mussels
arrange(td, site, desc(mussels))
```

**mutate**  
#mutate is similar to td$newcolumnname <- some_vectorized_operation, but you can create multiple columns at once
```{r}
mutate(td, musselsx2=mussels*2, logbarnacles=log(barnacles), snailplusmusselsx2=snails+musselsx2)

mutate(td, across(4:6, ~.x*2))
tide1 <-mutate(td, across(mussels:barnacles, as.character))

mutate(td, dominant_spp = ifelse(mussels > barnacles, "mussels","barnacles"))

mutate(td, dominant_spp = ifelse(mussels > barnacles, "mussels",
                                 ifelse(mussels==barnacles,"equal","barnacles")))



mutate(td, forage_hab= ifelse(tide.height=="high" & mussels > barnacles, "good forage", "suboptimal forage"))

```



**find & replace with gsub** 
```{r}
# edit a column using mutate
# gsub - a variant of grep, is like find replace and can be used 


```

**group and summarize data by column**
summarize is similar to aggregate() but you can create multiple aggregated columns at once
```{r}
summarize(td, mean=mean(mussels), n=n(), sd=sd(mussels), se=sd/sqrt(n))

td.grouped <- group_by(td, site) #grouping here by site
summarize(td.grouped, mean=mean(mussels), n=n(), sd=sd(mussels), se=sd/sqrt(n))

td.grouped2 <- group_by(td, site, tide.height)
summarize(td.grouped2, mean=mean(mussels), n=n(), sd=sd(mussels), se=sd/sqrt(n))

summarize(td.grouped2, mean=mean(mussels), n=n(), sd=sd(mussels), se=sd/sqrt(n), .groups="keep") #same grouping structure as the data

summarize_at(td.grouped2, vars(mussels:barnacles), list(m=mean, sd=sd))


mutate(td, mussels_scale=scale(mussels))

#scale data within sites
mutate(td.grouped, mussels_scale=scale(mussels))

```

**piping**
#The %>% operator in dplyr allows you to pass the output of a function as the 
first argument (input dataframe) for the next function, e.g.
**filter(tidepool_dat, Site=="A") is the same as tidepool_dat %>% filter(Site=="A")**
It eliminates the need to store lots of intermediates when manipulating data.
```{r}

data.a <- group_by(td, site, tide.height)
data.b <- summarize(data.a, mean=mean(mussels), n=n())
data.c <- filter(data.b, tide.height=="high")
data.d <- arrange(data.c, mean)
data.d

data2 <- td %>% 
 group_by(site, tide.height)%>%
  summarize(mean=mean(mussels), n=n()) %>%
 filter(tide.height=="high") %>%
 arrange(mean)
 data2

```



**unite, separate, rename**  
to make new variables. 
```{r}
# unite
td2 <-td %>% unite("siterep", site, replicate, sep="_", remove=TRUE)



# separate & rename (with backwards pipe!)
td2 <-td2 %>% separate(siterep, into=c("site", "replicate"), remove=T) %>% rename(Site=site)

```

**join**
joining tables (similar to merge())
join functions are useful when working with relational tables/databases.
```{r}
## create an example dataset ###
#filter the dataset to include sites A & C only. Then add a column called "df" which signifies the dataframe it belongs to.
tidepool_AC <-filter(td, site !="B")%>%mutate(df="tidepool_AC")

#filter the dataset to include sites B & C only.
tidepool_BC <-filter(td, site %in% c("B","C"))%>% mutate(df2 = "tidepool_BC")

#### LEFT JOIN ### - the dataframe on the left will overwrite the dataframe on the right.
tidepool2 <- left_join(tidepool_AC, tidepool_BC)

tidepool2 <-left_join(tidepool_AC, tidepool_BC, by=c("site"))

### RIGHT JOIN ### - the dataframe on the right will overwrite the dataframe on the left. 
tidepool2<-right_join(tidepool_AC, tidepool_BC)

## FULL JOIN ### - neither dataframes will overight. columns that have duplicate rows will retain both. 
tidepool2<-full_join(tidepool_AC, tidepool_BC)

```

**pivot** 
Convert between long format and wide format data frames. 
These functions can also be piped with dplyr functions.
```{r}

td2





####pivot_longer### 
#  rather than have separate columns for mussels, snails, barnacles, we can put all the numeric values into a column called 'Adundance' and create a new categorical column called 'Species', which has levels mussels, snails, barnacles. This is a conversion to long format.
td_long <-pivot_longer(td, names_to="Species", values_to="Abundance", mussels:barnacles)%>%as.data.frame()



####pivot_wider#####
# let's change td_long back to wide format. 
td_wide <-pivot_wider(td_long, names_from="Species", values_from="Abundance")

```
dplyr/tidyr functions have a tendency to convert things to 'tibbles' without asking, which sometimes aren't compatible with other code you've written for data frames. To avoid this, I will end piped sequences with %>% as.data.frame()

grouping also sometimes screws things up down the line, so it's often not a bad idea to %>% ungroup() after you group and perform operations.


**working with factor levels** from the "forcats" package
```{r}
td$site <-fct_recode(td$site, "1"="A", "2"="B", "3"="C")

td <-mutate(td, site=fct_recode(site, "A"="1", "B"="2", "C"="3"))

# combine sites B and C into one site called B
tidepool2 <-mutate(td, site=fct_collapse(site, "B"=c("B","C")))


```
